Solar wind prediction for the Parker Solar Probe orbit

### 2017-06-01 paperMVVB/sections/abstract.tex
introduced Parker Solar Probe, PSP

\title{Solar wind prediction for the Solar Probe Plus orbit}
\subtitle{Near-Sun extrapolations derived from an empirical solar wind model based on Helios and OMNI observations}

@@

\abstract
{In view of the planned near-Sun spacecraft mission Solar~Probe~Plus (SPP) the solar wind environment for its mission duration (2018--2025) and down to its intended closest perihelion (\num{9.86}~solar radii) is extrapolated using in situ data. This yet uncharted region is of special interest because the critical Alfvénic and sonic surfaces are believed to lie within and thus still part of the solar wind acceleration as well.}	%context
{We present an empirical solar wind model for the inner heliosphere which is derived from Helios and OMNI in situ data. The space probes Helios~1 and Helios~2 flew in the 1970s and observed solar wind in the ecliptic within heliocentric distances of \SIrange{0.29}{0.98}{\au}. The OMNI data set consists of multi-spacecraft intercalibrated in situ data from \SI{1}{\au}. This model is used together with sunspot number predictions to estimate the frequency distributions of major solar wind parameters SPP will encounter on its mission.}	%aims
{The model covers the solar wind's magnetic field strength and its plasma parameters proton density, velocity and temperature.
Their individual frequency distributions are represented with lognormal functions. In addition, we also consider the velocity distribution's bi-componental shape, consisting of a slower and a faster part. The model accounts for solar activity and for solar distance dependency by shifting of these lognormal distributions. We compile functional relations to solar activity by correlating and fitting the frequency distributions with the sunspot number (SSN), using almost five solar cycles of OMNI data. Further, based on the combined data set from both Helios probes, the parameters' frequency distributions are fitted over solar distance to obtain exponential dependencies. Combining both the solar cycle and the solar distance relations, we get a simple dynamical solar wind model for the inner heliosphere, confined to the ecliptic.}	%methods
{The inclusion of SSN predictions and the extrapolation to the SPP perihelion region enables us to estimate the solar wind environment at SPP's planned orbital positions during its mission time.}	%results
{}	%conclusions
%\SI{0.0459}{\au}

@@

%\titlerunning{Solar wind extrapolation to SPP orbit}
%\authorrunning{Venzmer \& Bothmer}

%extrapolated ranges \SI{95.45}{\percent} of the solar wind parameter values lie within. The ranges are for the magnetic field strength \SIrange{522}{2725}{nT}, for the velocity \SIrange{209.4}{506}{\km\per\s}, for the density \SIrange[scientific-notation=fixed, fixed-exponent=3, range-units=brackets]{1.319e3}{1.72e4}{\per\cm\cubed} and for the temperature \SIrange[scientific-notation=fixed, fixed-exponent=6, range-units=brackets]{4.18e5}{1.59e7}{\K}.
%These ranges are consistent with... and differ with...

>>>>

\title{Solar wind prediction for the Parker Solar Probe orbit}
\subtitle{Near-Sun extrapolations derived from an empirical solar wind model based on Helios and OMNI observations}

@@

\abstract
{In view of the planned near-Sun spacecraft mission Parker~Solar~Probe (PSP) (formerly Solar~Probe~Plus) the solar wind environment for its mission duration (2018--2025) and down to its intended closest perihelion (\num{9.86}~solar radii) is extrapolated using in situ data. This yet uncharted region is of special interest because the critical Alfvénic and sonic surfaces are believed to lie within and thus still part of the solar wind acceleration as well.}	%context
{We present an empirical solar wind model for the inner heliosphere which is derived from Helios and OMNI in situ data. The space probes Helios~1 and Helios~2 flew in the 1970s and observed solar wind in the ecliptic within heliocentric distances of \SIrange{0.29}{0.98}{\au}. The OMNI data set consists of multi-spacecraft intercalibrated in situ data from \SI{1}{\au}. This model is used together with sunspot number predictions to estimate the frequency distributions of major solar wind parameters PSP will encounter on its mission.}	%aims
{The model covers the solar wind's magnetic field strength and its plasma parameters proton density, velocity and temperature.
Their individual frequency distributions are represented with lognormal functions. In addition, we also consider the velocity distribution's bi-componental shape, consisting of a slower and a faster part. The model accounts for solar activity and for solar distance dependency by shifting of these lognormal distributions. We compile functional relations to solar activity by correlating and fitting the frequency distributions with the sunspot number (SSN), using almost five solar cycles of OMNI data. Further, based on the combined data set from both Helios probes, the parameters' frequency distributions are fitted over solar distance to obtain exponential dependencies. Combining both the solar cycle and the solar distance relations, we get a simple dynamical solar wind model for the inner heliosphere, confined to the ecliptic.}	%methods
{The inclusion of SSN predictions and the extrapolation to the PSP perihelion region enables us to estimate the solar wind environment at PSP's planned orbital positions during its mission time. The estimated solar wind values at PSP's nearest perihelion vary up to +-XX~\%, arising only from differing amplitude assumptions for the next solar cycle.}	%results
{}	%conclusions

@@

%\titlerunning{Solar wind extrapolation to PSP orbit}
%\authorrunning{Venzmer \& Bothmer}

%\SI{0.0459}{\au}
%extrapolated ranges \SI{95.45}{\percent} of the solar wind parameter values lie within. The ranges are for the magnetic field strength \SIrange{522}{2725}{nT}, for the velocity \SIrange{209.4}{506}{\km\per\s}, for the density \SIrange[scientific-notation=fixed, fixed-exponent=3, range-units=brackets]{1.319e3}{1.72e4}{\per\cm\cubed} and for the temperature \SIrange[scientific-notation=fixed, fixed-exponent=6, range-units=brackets]{4.18e5}{1.59e7}{\K}.
%These ranges are consistent with... and differ with...

### 2017-06-01 paperMVVB/sections/content.tex
introduced Parker Solar Probe, PSP

%in situ spacecraft
Several space missions flew away from Earth's orbit to measure in situ solar wind at different heliocentric distances (e.g. the probes Helios, Voyager, Ulysses...). Voyager~1 is the farthest probe out of the solar system, now covering more than 137~au (cite?). Ulysses was the single probe orbiting the Sun outside the ecliptic and retrieving solar wind measurements from the poles of the Sun. So far, Helios~2 made the nearest in situ solar wind measurements with \SI{0.29}{\au}, followed closely by Helios~1 with \SI{0.31}{\au}. The new mission Solar~Probe~Plus~(SPP), whose planned launch date is in mid 2018, aims to reach further down into the Sun's atmosphere.

%Solar Probe Plus mission
The main part of the SPP mission's goals is to gain insight into the coronal heating and solar wind acceleration processes. As these processes occur in the near-Sun region of the corona, SPP will have its closest perihelion at 9.86~solar radii (0.0459~au) \citep{Fox2015}. It will be the first spacecraft flying and measuring in situ solar wind this close to the Sun. Its planned observations range from the perihelion up to \SI{0.25}{\au}, leaving the outer part for data transfer, as \citet{Fox2015} state in their description of SPP's mission design.\\

@@

%solar wind parameters
The goal of this paper is to estimate the solar wind environment SPP will probe during its orbits. We regard the solar wind primarily as a proton plasma, because protons make up most of it. The average helium abundance is only about \SI{4.5}{\percent} and in slow wind at solar cycle minimum even less than \SI{2}{\percent} \citep{Feldman1978,Schwenn1983,Kasper2012}. Neglecting heavier ions, the electron and mass density can be derived accordingly.

@@

%frequency distribution data
One cannot know which specific solar wind type or structure SPP will encounter at a given point in time, so we extrapolate the parameters' probability distributions from existing solar wind measurements.

@@

This general model is then fed with a SSN prediction and extrapolated to SPP's planned orbital positions.\\

@@

%crossing distances
The next step is to fit the bulk of the solar wind parameters with lognormal functions. At all considered solar distances the mean of the three plasma parameters is larger than their median (Fig.~\ref{fig:radial_fit_4_thesis_light_skip_pdfcairo_plot}).
\begin{figure*}
	\includegraphics[width=18cm]{figures/radial_fit_4_thesis_light_skip_pdfcairo_plot.pdf}
	\caption{Helios hourly data plots of the four solar wind parameters over solar distance. The mean and median per \SI{0.01}{\au} data bin and their fit curves are plotted as well. The Helios data has a native distance resolution of \SI{0.01}{\au}. To make the abundance visible in these plots, we added a random distance value of up to \SI{+-0.005}{\au}.}
	\label{fig:radial_fit_4_thesis_light_skip_pdfcairo_plot}
\end{figure*}
However, the magnetic field strength's mean crosses the median at \SI{0.339}{\au} (Table~\ref{tab:mean_median_fit_parameter}) and is lower at smaller distances. At the crossing point and below the magnetic field strength can therefore not be described anymore with a lognormal function. For an extrapolation to the SPP perihelion the same happens for the temperature at \SI{0.082}{\au}.

@@

\section{Model extrapolation to SPP orbital time and position}
To estimate the solar wind environment at SPP's planned orbital positions during its mission time, SSN predictions are included into the general solar wind model and extrapolations to the SPP perihelion region are performed.\\

\subsection{SSN prediction for SPP mission time}
%SSN prediction
For the SSN prediction there are several sources available. The SIDC provides 12-month SSN forecasts\footnote{\url{http://sidc.be/silso/forecasts}} obtained from different methods (e.g. Kalman filter combined method). The Space Weather Prediction Center's (SWPC) prediction follows a consensus of the Solar~Cycle~24 Prediction~Panel\footnote{\url{http://www.swpc.noaa.gov/products/solar-cycle-progression}} (until end of 2019).

For the prediction of the next solar cycle~25 we simply assume a course similar to the last and thus we shift the last cycle by 11~years. Additionally we consider the two alternatives of half and twice its amplitude. The SSN for SPP's first perihelion will be small---certainly below 20, whereas its nearest perihelia, which commence at the height of cycle~25, will have as of now almost unpredictable SSN amplitudes (see Fig.~\ref{fig:SPP_orbit_predicted_SSN_overview_d_plot})\\
\begin{figure}
	\resizebox{\hsize}{!}{\includegraphics{figures/SPP_orbit_predicted_SSN_overview_d_plot.pdf}}
	\caption{SPP's solar distance during its mission time (top). Consecutive Venus flybys bring its perihelia nearer to the Sun. Actual and predicted SSN (bottom), i.e. SIDC 13-month smoothed monthly actual SSN, SIDC prediction, SWPC prediction and simply by 11~years shifted SSN from previous cycle~24, together with two alternative trends of half and twice its amplitude.}
	\label{fig:SPP_orbit_predicted_SSN_overview_d_plot}
\end{figure}

@@

\subsection{Near-Sun extrapolation for SPP orbit}

%SPP orbit
SPP orbit description, seven Venus flybys, within Venus' orbital plane? (\SI{3.86}{\degree} to Sun's equator), \citep{Fox2015}, (see Fig.~\ref{fig:SPP_orbit_predicted_SSN_overview_d_plot})\\

@@

The near-Sun (SPP perihelion) solar wind velocity is expected to be slower than our model's estimates, because the position of the source (Alfvénic critical) surface is predicted to lie between 15--30~Rs (Schatten1969, Sittler1999, Exarhos2000, Katsikas2010, Goelzer2014; choose references...), up to which the solar wind is believed to be accelerated.\\

@@

\subsection{SPP solar wind environment estimation}
Implementing the orbital distance data and predicted SSN for the mission time we get SPP's estimated solar wind environment $W'''(x,ssn,r)$.\\
The zoom into the first and the nearest perihelia show which solar wind parameter magnitudes can be expected there (Figs.~\ref{fig:SPP_perihelia_prediction_d_plot} and \ref{fig:SPP_perihelia_prediction_nearest_d_plot}).\\
\begin{figure}
	\resizebox{\hsize}{!}{\includegraphics{figures/SPP_perihelia_prediction_d_plot.pdf}}
	\caption{Estimated solar wind parameter medians (black) and their error bands (grey) during SPP's first perihelion in 2018 at about \SI{0.16}{\au}. For the velocity the SSN independent slow and fast parts are plotted.}
	\label{fig:SPP_perihelia_prediction_d_plot}
\end{figure}
\begin{figure}
	\resizebox{\hsize}{!}{\includegraphics{figures/SPP_perihelia_prediction_nearest_d_plot.pdf}}
	\caption{Estimated solar wind parameter medians (black) and their error bands (grey) during SPP's nearest perihelion in 2024 at \SI{0.0459}{\au}. For the velocity the SSN independent slow and fast parts are plotted.}
	\label{fig:SPP_perihelia_prediction_nearest_d_plot}
\end{figure}

>>>>

%in situ spacecraft
Several space missions flew away from Earth's orbit to measure in situ solar wind at different heliocentric distances (e.g. the probes Helios, Voyager, Ulysses...). Voyager~1 is the farthest probe out of the solar system, now covering more than 137~au (cite?). Ulysses was the single probe orbiting the Sun outside the ecliptic and retrieving solar wind measurements from the poles of the Sun. So far, Helios~2 made the nearest in situ solar wind measurements with \SI{0.29}{\au}, followed closely by Helios~1 with \SI{0.31}{\au}. The new mission Parker~Solar~Probe (PSP), whose planned launch date is in mid 2018, aims to reach further down into the Sun's atmosphere.

%Solar Probe Plus mission
The main part of the PSP mission's goals is to gain insight into the coronal heating and solar wind acceleration processes. As these processes occur in the near-Sun region of the corona, PSP will have its closest perihelion at 9.86~solar radii (0.0459~au) \citep{Fox2015}. It will be the first spacecraft flying and measuring in situ solar wind this close to the Sun. Its planned observations range from the perihelion up to \SI{0.25}{\au}, leaving the outer part for data transfer, as \citet{Fox2015} state in their description of PSP's mission design.\\

@@

%solar wind parameters
The goal of this paper is to estimate the solar wind environment PSP will probe during its orbits. We regard the solar wind primarily as a proton plasma, because protons make up most of it. The average helium abundance is only about \SI{4.5}{\percent} and in slow wind at solar cycle minimum even less than \SI{2}{\percent} \citep{Feldman1978,Schwenn1983,Kasper2012}. Neglecting heavier ions, the electron and mass density can be derived accordingly.

@@

%frequency distribution data
One cannot know which specific solar wind type or structure PSP will encounter at a given point in time, so we extrapolate the parameters' probability distributions from existing solar wind measurements.

@@

This general model is then fed with a SSN prediction and extrapolated to PSP's planned orbital positions.\\

@@

%crossing distances
The next step is to fit the bulk of the solar wind parameters with lognormal functions. At all considered solar distances the mean of the three plasma parameters is larger than their median (Fig.~\ref{fig:radial_fit_4_thesis_light_skip_pdfcairo_plot}).
\begin{figure*}
	\includegraphics[width=18cm]{figures/radial_fit_4_thesis_light_skip_pdfcairo_plot.pdf}
	\caption{Helios hourly data plots of the four solar wind parameters over solar distance. The mean and median per \SI{0.01}{\au} data bin and their fit curves are plotted as well. The Helios data has a native distance resolution of \SI{0.01}{\au}. To make the abundance visible in these plots, we added a random distance value of up to \SI{+-0.005}{\au}.}
	\label{fig:radial_fit_4_thesis_light_skip_pdfcairo_plot}
\end{figure*}
However, the magnetic field strength's mean crosses the median at \SI{0.339}{\au} (Table~\ref{tab:mean_median_fit_parameter}) and is lower at smaller distances. At the crossing point and below the magnetic field strength can therefore not be described anymore with a lognormal function. For an extrapolation to the PSP perihelion the same happens for the temperature at \SI{0.082}{\au}.

@@

\section{Model extrapolation to PSP orbital time and position}
To estimate the solar wind environment at PSP's planned orbital positions during its mission time, SSN predictions are included into the general solar wind model and extrapolations to the PSP perihelion region are performed.\\

\subsection{SSN prediction for PSP mission time}
%SSN prediction
For the SSN prediction there are several sources available. The SIDC provides 12-month SSN forecasts\footnote{\url{http://sidc.be/silso/forecasts}} obtained from different methods (e.g. Kalman filter combined method). The Space Weather Prediction Center's (SWPC) prediction follows a consensus of the Solar~Cycle~24 Prediction~Panel\footnote{\url{http://www.swpc.noaa.gov/products/solar-cycle-progression}} (until end of 2019).

For the prediction of the next solar cycle~25 we simply assume a course similar to the last and thus we shift the last cycle by 11~years. Additionally we consider the two alternatives of half and twice its amplitude. The SSN for PSP's first perihelion will be small---certainly below 20, whereas its nearest perihelia, which commence at the height of cycle~25, will have as of now almost unpredictable SSN amplitudes (see Fig.~\ref{fig:SPP_orbit_predicted_SSN_overview_d_plot})\\
\begin{figure}
	\resizebox{\hsize}{!}{\includegraphics{figures/SPP_orbit_predicted_SSN_overview_d_plot.pdf}}
	\caption{PSP's solar distance during its mission time (top). Consecutive Venus flybys bring its perihelia nearer to the Sun. Actual and predicted SSN (bottom), i.e. SIDC 13-month smoothed monthly actual SSN, SIDC prediction, SWPC prediction and simply by 11~years shifted SSN from previous cycle~24, together with two alternative trends of half and twice its amplitude.}
	\label{fig:SPP_orbit_predicted_SSN_overview_d_plot}
\end{figure}

@@

\subsection{Near-Sun extrapolation for PSP orbit}

%PSP orbit
PSP orbit description, seven Venus flybys, within Venus' orbital plane? (\SI{3.86}{\degree} to Sun's equator), \citep{Fox2015}, (see Fig.~\ref{fig:SPP_orbit_predicted_SSN_overview_d_plot})\\

@@

The near-Sun (PSP perihelion) solar wind velocity is expected to be slower than our model's estimates, because the position of the source (Alfvénic critical) surface is predicted to lie between 15--30~Rs (Schatten1969, Sittler1999, Exarhos2000, Katsikas2010, Goelzer2014; choose references...), up to which the solar wind is believed to be accelerated.\\

@@

\subsection{PSP solar wind environment estimation}
Implementing the orbital distance data and predicted SSN for the mission time we get PSP's estimated solar wind environment $W'''(x,ssn,r)$.\\
The zoom into the first and the nearest perihelia show which solar wind parameter magnitudes can be expected there (Figs.~\ref{fig:SPP_perihelia_prediction_d_plot} and \ref{fig:SPP_perihelia_prediction_nearest_d_plot}).\\
\begin{figure}
	\resizebox{\hsize}{!}{\includegraphics{figures/SPP_perihelia_prediction_d_plot.pdf}}
	\caption{Estimated solar wind parameter medians (black) and their error bands (grey) during PSP's first perihelion in 2018 at about \SI{0.16}{\au}. For the velocity the SSN independent slow and fast parts are plotted.}
	\label{fig:SPP_perihelia_prediction_d_plot}
\end{figure}
\begin{figure}
	\resizebox{\hsize}{!}{\includegraphics{figures/SPP_perihelia_prediction_nearest_d_plot.pdf}}
	\caption{Estimated solar wind parameter medians (black) and their error bands (grey) during PSP's nearest perihelion in 2024 at \SI{0.0459}{\au}. For the velocity the SSN independent slow and fast parts are plotted.}
	\label{fig:SPP_perihelia_prediction_nearest_d_plot}
\end{figure}
